Vagueness and imprecise imitation in signaling games

### 2016-02-23 paper/version_05/paper.tex
some improvements on intro, references added for semantic maps

But there are other intriguing aspects about vagueness. One perplexing issue is how vagueness
could arise and be maintained in the first place. This is an apparent puzzle for functionalist
accounts that maintain that concepts and linguistic meanings evolved driven towards
efficiency. Since the existence of unclear borderline cases seems to entail inefficiency of
categorization or communication, the challenge, succinctly put forward by
\citet{Lipman2009:Why-is-Language}, is to explain two things at the same time: (i) how
vagueness can persist under evolutionary pressure to be optimally discriminative, (ii) without
jeopardizing the possibility of evolving, learning and communicating with a meaningful
language. A number of authors have consequently tried to explain why vagueness evolved as
something that is itself useful
\citep[e.g.][]{Jaegherde-Jaegher2003:A-Game-Theoreti,Deemter2009:Utility-and-Lan,BlumeBoard2013:Intentional-Vag}.
Others have argued that vagueness is a natural byproduct of limitations in information
processing \citep[e.g.][]{FrankeJager2010:Vagueness-Signa} or of generalization in low-level
learning strategies \citep[e.g.][]{OConnor2013:The-Evolution-o}. This paper contributes to the
latter line of thought. We believe that vagueness in language and thought may have many
reasons, not just one. We focus here on one possibility that has not received due attention,
namely imprecise perception of similar stimuli and imprecise realization of similar
responses.

Our conceptual starting point is the widely shared conviction that language is shaped by at
least two forces, which may, on occasion, pull in opposite direction. On the one hand, there is
functional pressure towards efficient communication. On the other hand, there is systematic
error, noise or imprecision in the transmission of linguistic behavior, knowledge or traits. As
an example for the latter, consider a child learning syntactic rules from a parent
generation. The child must infer these unobservable rules from observable speech. Inductive
biases of the learner may influence which syntactic rules are likely to be inferred from
(finite) parental input. Over the course of many generations, the effects of such transmission
biases can, in a manner of speaking, ``accumulate'' and lead to surprising results, such as the
evolution of compositional form-meaning mappings or the use of regular recursive syntactic
structure. This can happen when the effects of tranmission biases are isolated, as in iterated
learning models
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,GriffithsKalish2007:Language-Evolut,KirbyGriffith2014:Iterated-Learni},
or when they interact with functional pressure towards efficient communication, e.g., as
formalized in the replicator mutator dynamic
\citep[e.g.][]{NowakPlotkin2000:The-Evolution-o,NowakKomarova2001:Evolution-of-Un}.

The emphasis of previous models that studied the effects of transmission infidelity on the
evolution of language has been on inductive biases and the systematicity, compression and
regularization that they can introduce. Here, we would like to show that transmission noise of
a different kind can lead to regularization as well and also give rise to vague
meaning. Towards this end, we formalize the expected change in the behavior of a population of
agents that try to imitate other agents' signaling behavior. We assume, however, that both
realization of behavior, as well as observability of others' behavior, is molested by noise, in
particular the probabilistic confusability of similar stimuli. The resulting population-level
dynamic generalizes the replicator dynamic \citep{TaylorJonker1978:Evolutionary-St} in its
interpretation as a cultural evolutionary dynamic based on imitation
\citep[e.g.][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and}. The inclusion of
confusability of stimuli does not undermine the possibility of evolving communicative signaling
behavior. Instead, it leads to the evolution of vague meanings. It moreover unifies and
regularizes evolutionary outcomes, making it appear as if agents were applying inductive biases
or generalizing over partial observations, when in truth this is the sole effect of confusion
of similar stimuli.

@@

Similarity-maximizing (short: sim-max) games are a variation of
Lewis' games where the receiver's actions are equated with the state
space (one can think of the actions as choosing states) and different states are allowed to be more or less
similar to one another. While Lewis' games treated communicative
success as a matter of black and white, sim-max games allow for a
gradient notion: the more similar the receiver's interpretation is to
the actual state, the better. Signaling games with utility-relevant
similarities in the state space are fairly standard in economics
\citep[e.g.][]{Spence1973:Job-market-sign,CrawfordSobel1982:Strategic-Infor},
but have received particular attention in a more philosophical context
for reasons that will become clear presently.

@@

It is intuitive to think that linguistic and conceptual categories are orderly in such a manner.
For example, if you consider two people, one with a height of 2m and another with a height of 2.2m, both `tall', it would be difficult to defend considering a person with a height of 2.1m `short'.
Another example is that of color categorization.
The World Color Survey project~\citep{CookKay2005:The-World-Color,WCS} collected color naming data for 110 unwritten languages of 45 language families.
For the great majority of these languages a pattern can be observed: basic color terms are by and large convex~\citep{Regier07,Jager2010}.
It is premature to argue that these observations can be extended to \emph{all} cases of categorization.
However, for the cases where it does apply, the result of \citet{JagerMetzger2011:Voronoi-Languag} is interesting because it demonstrates that signaling can impose this kind of orderly categories on a metric space without that being the ulterior purpose of it all.%
\footnote{For the concrete case of color categorization, see also \citet{JagerRooijvan-Rooij2007:Language-Struct} and \citet{Correia2015}}

>>>>

But there are other intriguing aspects about vagueness. One perplexing issue is how vagueness
could arise and be maintained in the first place. This is an apparent puzzle for functionalist
accounts that maintain that concepts and linguistic meanings evolved driven towards
efficiency. Since the existence of unclear borderline cases seems to entail inefficiency of
categorization or communication, the challenge, succinctly put forward by
\citet{Lipman2009:Why-is-Language}, is to explain two things at the same time: (i) how
vagueness can persist under evolutionary pressure to be optimally discriminative, (ii) without
jeopardizing the possibility of evolving, learning and communicating with a meaningful
language. A number of authors have consequently tried to explain why vagueness evolved as
something that is itself useful
\citep[e.g.][]{Jaegherde-Jaegher2003:A-Game-Theoreti,Deemter2009:Utility-and-Lan,BlumeBoard2013:Intentional-Vag}.
Others have argued that vagueness is a natural byproduct of limitations in information
processing \citep[e.g.][]{FrankeJager2010:Vagueness-Signa} or of generalization in low-level
learning strategies \citep[e.g.][]{OConnor2013:The-Evolution-o}. This paper contributes to the
latter line of thought.

We believe that vagueness in language and thought may have many reasons, not just one. We focus
here on one \emph{a priori} plausible reason for why vagueness is natural and pervasive. The
idea is that at least some meanings are vague, in part because of imprecision in the perception
of similar stimuli and in the realization of similar responses.\footnote{On this view,
  vagueness in language may be seen as a necessary sub-optimality due to limitations in another
  domain of cognition, first and foremost perception. The whole system, e.g., perception and
  language together, may be an almost optimal adaptation in a larger frame of reference, e.g.,
  if we take into account the metabolic costs for increased accuracy.} While this idea may be
\emph{prima facie} plausible, it is not immediately clear what the consequences of
confusability of stimuli are on a process of meaning evolution. In other words, in order to
address Lipman's challenge seriously, and not just hand-wave it away, it needs to be spelled
out how exactly confusability of stimuli could come into play in a process of meaning evolution
and how it could lead to vague but by-and-large informative signal meaning. The formal model
that this paper introduces does exactly that. And more. % We find that confusability of stimuli
% can regularize and systematize evolving meaning, and may thereby imply an advantageous
% side-effect that may counter some of the disadvantages that imprecise and vague communication
% may have. Here is some elaboration on this point.

Our conceptual starting point is the widely shared conviction that language is shaped by at
least two forces, which may, on occasion, pull in opposite direction. On the one hand, there is
functional pressure towards efficient communication. On the other hand, there is systematic
error, noise or imprecision in the transmission of linguistic behavior, knowledge or traits. As
an example for the latter, consider a child learning syntactic rules from a parent
generation. The child must infer these unobservable rules from observable speech. Inductive
biases may influence which syntactic rules are likely to be inferred from (finite) parental
input. Over the course of many generations, the effects of such biases can, in a manner of
speaking, ``accumulate'' and lead to surprising results, such as the evolution of compositional
form-meaning mappings or the use of regular recursive syntactic structure. This can happen when
the effects of tranmission biases are isolated, as in iterated learning models
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,GriffithsKalish2007:Language-Evolut,KirbyGriffith2014:Iterated-Learni},
or when they interact with functional pressure towards efficient communication, e.g., as
formalized in the replicator mutator dynamic
\citep[e.g.][]{NowakPlotkin2000:The-Evolution-o,NowakKomarova2001:Evolution-of-Un}.

The emphasis of previous models that studied the effects of transmission infidelity on the
evolution of language has been on inductive biases and the systematicity, compression and
regularization that they can introduce. Here, we would like to show that transmission noise of
a different kind can lead to regularization as well and also give rise to vague
meaning. Towards this end, we formalize the expected change in the behavior of a population of
agents that try to imitate other agents' signaling behavior. We assume, however, that both
realization of behavior, as well as observability of others' behavior, is molested by noise, in
particular the probabilistic confusability of similar stimuli. The resulting population-level
dynamic generalizes the replicator dynamic \citep{TaylorJonker1978:Evolutionary-St} in its
interpretation as a cultural evolutionary dynamic based on imitation
\citep[e.g.][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and}. The inclusion of
confusability of stimuli does not undermine the possibility of evolving communicative signaling
behavior. Instead, it leads to the evolution of vague meanings. It moreover accelerates the
emergence of communicative signaling because it unifies and regularizes evolutionary outcomes,
making it appear as if agents were applying inductive biases or generalizing over partial
observations, when in truth this is the sole effect of confusion of similar stimuli.

@@

Similarity-maximizing (short: sim-max) games are a variation of Lewis' games where the
receiver's actions are equated with the state space (one can think of the actions as choosing
states) and different states are allowed to be more or less similar to one another. While
Lewis' games treated communicative success as a matter of black and white, sim-max games allow
for shades of gray: the more similar the receiver's interpretation is to the actual state, the
better. Signaling games with utility-relevant similarities in the state space are fairly
standard in economics
\citep[e.g.][]{Spence1973:Job-market-sign,CrawfordSobel1982:Strategic-Infor}, but have received
particular attention in a more philosophical context for reasons that will become clear
presently.

@@

It is intuitive to think that linguistic and conceptual categories are orderly in such a
manner.  For example, if you consider two people, one with a height of 2m and another with a
height of 2.2m, both `tall', it would be difficult to defend considering a person with a height
of 2.1m `short'.  Another example is that of color categorization. The World Color Survey
project~\citep{CookKay2005:The-World-Color,WCS} collected color naming data for 110 unwritten
languages of 45 language families.  For the great majority of these languages a pattern can be
observed: basic color terms are by and large convex~\citep{Regier07,Jager2010}. A prominent
school of thought in comparative linguistics also assumes that more abstract conceptual domains
---e.g., spatial-topological relations, temporal reference, or the meanings expressible by
indefinite pronouns--- are preferably carved up by the languages of the world in such a way
that meanings are connected regions on a so-called \emph{semantic map}
\citep[e.g.][]{Croft2003:Typology-and-Un,Haspelmath2003:The-geometry-of,LevinsonMeira2003:Natural-concept}. It
is premature to argue that these observations can be extended to \emph{all} cases of
categorization. However, for the cases where it does apply, the result of
\citet{JagerMetzger2011:Voronoi-Languag} is interesting because it demonstrates that signaling
can impose this kind of orderly categories on a metric space without that being the ulterior
purpose of it all.%
\footnote{For the concrete case of color categorization, see also
  \citet{JagerRooijvan-Rooij2007:Language-Struct} and \citet{Correia2015}}
